---
title: "5.1.1 Axe cadre bâti - Répartition des logements"
author: "Curbcut"
output: html_document
---

```{r setup_5_1_1, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(tidyverse)
library(gt)
qs::qload(here::here("data", "section_5_1_1.qsm"))
```

# 5.1.1.1 Répartition des logements selon le typologie

```{r housing_type_figures}

single_shares <- 
  housing |> 
  sf::st_drop_geometry() |> 
  summarize(total = sum(type_total, na.rm = TRUE),
            single = sum(type_single, na.rm = TRUE),.by = year) |> 
  mutate(pct = single / total) |> 
  pull(pct) |> 
  scales::percent(0.1)

```

_Data source: Canadian Census_

From 2001 to 2021, the total number of privately occupied dwelling units in the City of Laval has grown at an average rate of 1.32% per year, almost identical to the Canadian average (1.31%). In all five census periods, single-detached dwellings have represented a plurality of all dwellings, although this share has been declining, from `r single_shares[1]` in 2001 to `r single_shares[5]` in 2021. This decline has been met with an increase in low-rise apartments in the early 2000s and then an increase in medium-rise and high-rise apartments in the last census period. In both proportional and absolute terms, semi-detached and duplex housing types have experienced steady declines, while the count of row houses has grown in absolute terms although it has remained mostly constant as a share of all dwelling units.

```{r plot_5_1_1_1_facet}
plot_5_1_1_1_facet
```

```{r table_5_1_1_1}
table_5_1_1_1
```

The share of all private dwelling units which are single-family homes is lowest in the southern part of the island; in the Marigot, Souvenir-Labelle, Concorde-Bois-de-Boulogne, L'Abord-à-Plouffe, Laval-des-Rapides, and Renaud electoral districts, single-family homes make up fewer than one in four total occupied private dwellings. This proportion is far higher in the west and north-east portions of the city; in the Laval-les-Îles, Auteuil, L'Orée-des-Bois, Saint-François, Sainte-Dorothée, Marc-Aurèle-Fortin, and Val-des-Arbres electoral districts more than six homes in ten are single-detached.

```{r map_5_1_1_1_single}
map_5_1_1_1_single
```

# 5.1.1.2 Répartition des logements selon le mode d'occupation

```{r housing_tenure_figures}

owner_shares <- 
  housing |> 
  sf::st_drop_geometry() |> 
  summarize(total = sum(tenure_total, na.rm = TRUE),
            owner = sum(tenure_owner, na.rm = TRUE),.by = year) |> 
  mutate(pct = owner / total) |> 
  pull(pct) |> 
  scales::percent(0.1)

```

_Data source: Canadian Census_

Homeownership rates have been very steady in Laval since 2001; the share of total occupied private dwellings occupied by owner households was `r owner_shares[1]` in 2001 and was also `r owner_shares[5]` in 2021, and in the intervening years never rose higher than `r owner_shares[3]`. Put differently, although the last ten years have seen a steady uptick in the total proportion of the city's housing stock occupied by renters (corresponding to an increase in apartment construction discussed in section 6, below), this uptick has simply cancelled out a similar decline in the renter share households which occurred from 2001 to 2011. 
```{r plot_5_1_1_2_facet}
plot_5_1_1_2_facet
```

```{r table_5_1_1_2}
table_5_1_1_2
```

The spatial patterns of owner-occupied dwellings in Laval closely mirror the patterns of single-family homes analyzed in the previous section; southern sectors have low rates (< 50%) of owner-occupied dwellings, while the west and north-eastern areas of the island have high rates (> 80%).

```{r map_5_1_1_2_owner}
map_5_1_1_2_owner
```

# 5.1.1.3 Répartition des logements selon le nombre de chambres à coucher

_Data source: Canadian Census_

# 5.1.1.4 Répartition des logements selon l'âge du cadre bâti

_Data source: Canadian Census_

# 5.1.1.5 Répartition des logements selon le type de parc : privé lucratif, non lucratif (privé ou public)



