---
title: "5.1.1 Axe cadre bâti - Répartition des logements"
author: "Curbcut"
output: html_document
---

```{r setup_5_1_1, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(tidyverse)
library(gt)
qs::qload(here::here("data", "section_5_1_1.qsm"))
```

# 5.1.1.1 Répartition des logements selon le typologie

```{r housing_type_figures}

single_shares <- 
  housing |> 
  sf::st_drop_geometry() |> 
  summarize(total = sum(type_total, na.rm = TRUE),
            single = sum(type_single, na.rm = TRUE),.by = year) |> 
  mutate(pct = single / total) |> 
  pull(pct) |> 
  scales::percent(0.1)

```

_Data source: Canadian Census_

From 2001 to 2021, the total number of privately occupied dwelling units in the City of Laval has grown at an average rate of 1.33% per year, almost identical to the Canadian average (1.31%). In all five census periods, single-detached dwellings have represented a plurality of all dwellings, although this share has been declining, from `r single_shares[1]` in 2001 to `r single_shares[5]` in 2021. This decline has been met with an increase in low-rise apartments in the early 2000s and then an increase in medium-rise and high-rise apartments in the last census period. In both proportional and absolute terms, semi-detached and duplex housing types have experienced steady declines, while the count of row houses has grown in absolute terms although it has remained mostly constant as a share of all dwelling units.

```{r plot_5_1_1_1_facet}
plot_5_1_1_1_facet
```

```{r table_5_1_1_1}
table_5_1_1_1
```

The share of all private dwelling units which are single-family homes is lowest in the southern part of the island; in the Marigot, Souvenir-Labelle, Concorde-Bois-de-Boulogne, L'Abord-à-Plouffe, Laval-des-Rapides, and Renaud electoral districts, single-family homes make up fewer than one in four total occupied private dwellings. This proportion is far higher in the west and north-east portions of the city; in the Laval-les-Îles, Auteuil, L'Orée-des-Bois, Saint-François, Sainte-Dorothée, Marc-Aurèle-Fortin, and Val-des-Arbres electoral districts more than six homes in ten are single-detached.

```{r map_5_1_1_1_single}
map_5_1_1_1_single
```

# 5.1.1.2 Répartition des logements selon le mode d'occupation

```{r housing_tenure_figures}

owner_shares <- 
  housing |> 
  sf::st_drop_geometry() |> 
  summarize(total = sum(tenure_total, na.rm = TRUE),
            owner = sum(tenure_owner, na.rm = TRUE),.by = year) |> 
  mutate(pct = owner / total) |> 
  pull(pct) |> 
  scales::percent(0.1)

```

_Data source: Canadian Census_

Homeownership rates have been very steady in Laval since 2001; the share of total occupied private dwellings occupied by owner households was `r owner_shares[1]` in 2001 and was also `r owner_shares[5]` in 2021, and in the intervening years never rose higher than `r owner_shares[3]`. Put differently, although the last ten years have seen a steady uptick in the total proportion of the city's housing stock occupied by renters (corresponding to an increase in apartment construction discussed in section 6, below), this uptick has simply cancelled out a similar decline in the renter share households which occurred from 2001 to 2011. 
```{r plot_5_1_1_2_facet}
plot_5_1_1_2_facet
```

```{r table_5_1_1_2}
table_5_1_1_2
```

The spatial patterns of owner-occupied dwellings in Laval closely mirror the patterns of single-family homes analyzed in the previous section; southern sectors have low rates (< 50%) of owner-occupied dwellings, while the west and north-eastern areas of the island have high rates (> 80%).

```{r map_5_1_1_2_owner}
map_5_1_1_2_owner
```

# 5.1.1.3 Répartition des logements selon le nombre de chambres à coucher

_Data source: Canadian Census_

Although three-bedroom homes are the most common housing size in Laval, their share of the total housing stock has been declining steadily over the 2011-2021 period during which bedroom counts were measured by the Census. In fact, despite robust overall growth in housing units, as of 2021 there were fewer three-bedroom units than there were in 2011. All other bedroom counts saw absolute growth in unit counts from 2011-2021, but that growth was highest among the largest units: the share of total housing stock composed of units with four or more bedrooms grew by nearly a quarter.

```{r plot_5_1_1_3_facet}
plot_5_1_1_3_facet
```

```{r table_5_1_1_3}
table_5_1_1_3
```

Over the past ten years, a growing (albeit still small) proportion of dwelling units in the outlying areas of the city have four or more bedrooms. In the Val-des-Arbres, Saint-François, Auteuil, Laval-les-Îles, Marc-Aurèle-Fortin, Sainte-Dorothée, Fabreville, and Saint-Martin electoral districts, more than one quarter of homes had four or more bedrooms in 2021, while in 2011 it was only Saint-François and Val-des-Arbres. This trend points to the increasingly large houses which are common in new single-family construction.

```{r map_5_1_1_3_four}
map_5_1_1_3_four
```

# 5.1.1.4 Répartition des logements selon l'âge du cadre bâti

```{r construction_facts}
ages <- 
  housing |> 
  sf::st_drop_geometry() |> 
  filter(year == 2021) |> 
  select(NOM, age_total:age_2021) |> 
  mutate(age_2000 = age_1990 + age_2000,
         age_2021 = age_2005 + age_2010 + age_2015 + age_2021) |> 
  select(-age_1990, -age_2005, -age_2010, -age_2015) |> 
  summarize(across(age_total:age_2021, sum)) |> 
  mutate(across(age_1960:age_2021, \(x) x / age_total)) |> 
  mutate(across(age_1960:age_2021, scales::percent, 0.1))
```

_Data source: Canadian Census_

There is a relatively even distribution of building ages among private housing units in Laval, although recent decades are underrepresented in building ages compared to the postwar period. Comparing roughly 20-year construction period, `r ages$age_1960` of private occupied dwellings in 2021 were constructed in 1960 or earlier, `r ages$age_1980` were constructed between 1961 and 1980, `r ages$age_2000` were constructed between 1981 and 2000, and `r ages$age_2021` were constructed between 2001 and 2021. Put differently, the major period for housing construction in the city was the 1960s and 1970s, and subsequent decades have seen slowly decreasing rates of new housing arriving on the market, even in the face of the ongoing attrition of older units.

```{r plot_5_1_1_4}
plot_5_1_1_4
```

```{r table_5_1_1_4}
table_5_1_1_4
```

As the map indicates, there is a distinctive spatial pattern to different ages of residential buildings in the city. Housing units constructed in 1960 or earlier are only common along the south shore of Laval; in the electoral district of Duvernay-Pont-Viau they are nearly a majority of all private housing units, while elsewhere along the shore they are generally between 20 and 30 percent of all units. Meanwhile, in most of the city a plurality of all private housing units present in 2021 were constructed between 1961 and 1980, with majority concentrations in the Chomedey, Marc-Aurèle-Fortin and Marigot electoral districts. Housing construction in the 1980s and 1990s was concentrated mostly to the west and north of the island, while in the Renaud and Fabreville electoral districts more than one in ten private housing units in 2021 had been constructed in the previous five years.

```{r map_5_1_1_4}
map_5_1_1_4
```

# 5.1.1.5 Répartition des logements selon le type de parc : privé lucratif, non lucratif (privé ou public)

_Data source: Canadian Census_

```{r subsid_data}
subsid_data <- 
  housing |> 
  sf::st_drop_geometry() |> 
  filter(year >= 2011) |> 
  summarize(across(subsid_total:subsid, sum), .by = year) |> 
  mutate(pct = scales::percent(subsid / subsid_total, 0.1)) |> 
  mutate(subsid = scales::comma(subsid, 10))

can_subsid_pct <- scales::percent(can_subsid$subsid / 100, 0.1)
qc_subsid_pct <- scales::percent(qc_subsid$subsid / 100, 0.1)
```

There is only limited systematic information available about households living in non-market housing. In particular, for the last ten years the Canadian Census has identified tenant households who live in subsidized housing. According to this indicator, the share of the overall tenant population in Laval living in subsidized housing has been flat, from `r subsid_data$subsid[1]` (`r subsid_data$pct[1]`) in 2011 to `r subsid_data$subsid[3]` (`r subsid_data$pct[3]`) in 2021. This rate is quite a bit lower than both the Canadian (`r can_subsid_pct`) and Quebec (`r qc_subsid_pct`) rates.

```{r table_5_1_1_5}
table_5_1_1_5
```

Although the geography of subsidized housing has shifted somewhat over the last ten years, the electoral districts of Souvenir-Labelle and Renaud consistently have had the highest concentrations of subsidized tenant households, ranging between 10 and 20 percent of total tenant households between 2011 and 2021.

```{r map_5_1_1_5}
map_5_1_1_5
```


