---
title: "5.1.1 Axe cadre bâti - Répartition des logements"
author: "Curbcut"
output: html_document
---

```{r setup_5_1_1, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(tidyverse)
library(gt)
qs::qload(here::here("data", "section_5_1_1.qsm"))
```

# 5.1.1.1 Répartition des logements selon le typologie

```{r housing_figures}

single_shares <- 
  housing |> 
  st_drop_geometry() |> 
  summarize(total = sum(type_total, na.rm = TRUE),
            single = sum(type_single, na.rm = TRUE),.by = year) |> 
  mutate(pct = single / total) |> 
  pull(pct) |> 
  scales::percent(0.1)

```

_Data source: Canadian Census_

From 2001 to 2021, the total number of privately occupied dwelling units in the City of Laval has grown at an average rate of 1.32% per year, almost identical to the Canadian average (1.31%). In all five census periods, single-detached dwellings have represented a plurality of all dwellings, although this share has been declining, from `r single_shares[1]` in 2001 to `r single_shares[5]` in 2021. This decline has been met with an increase in low-rise apartments in the early 2000s and then an increase in medium-rise and high-rise apartments in the last census period. In both proportional and absolute terms, semi-detached and duplex housing types have experienced steady declines, while the count of row houses has grown in absolute terms although it has remained mostly constant as a share of all dwelling units.

```{r plot_5_1_1_1_facet}
plot_5_1_1_1_facet
```

```{r table_5_1_1_1}
table_5_1_1_1
```

The share of all private dwelling units which are single-family homes is lowest in the southern part of the island; in the Marigot, Souvenir-Labelle, Concorde-Bois-de-Boulogne, L'Abord-à-Plouffe, Laval-des-Rapides, and Renaud electoral districts, single-family homes make up fewer than one in four total occupied private dwellings. This proportion is far higher in the west and north-east portions of the city; in the Laval-les-Îles, Auteuil, L'Orée-des-Bois, Saint-François, Sainte-Dorothée, Marc-Aurèle-Fortin, and Val-des-Arbres electoral districts more than six homes in ten are single-detached.

```{r map_5_1_1_1_single}
map_5_1_1_1_single
```

# 5.1.1.2 Répartition des logements selon le mode d'occupation

# 5.1.1.3 Répartition des logements selon le nombre de chambres à coucher

# 5.1.1.4 Répartition des logements selon l'âge du cadre bâti

# 5.1.1.5 Répartition des logements selon le type de parc : privé lucratif, non lucratif (privé ou public)



