---
title: "6.1 À but lucratif (abordable, prix du marché)"
author: "Curbcut"
output: html_document
---

```{r setup_6_1, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
qs::qload(here::here("data", "section_6_1.qsm"))
library(tidyverse)
library(gt)
library(patchwork)

```

# 6.1.1 Développement résidentiel récent (répartition territoriale par type de logement et typologie)

```{r s_6_1_1, include = FALSE}

# Min and max 5-year avg
min_max_vals <- 
  completions_by_type |> 
  filter(is.na(zone)) |>
  filter(type == "All") |> 
  mutate(value_trend = slider::slide_dbl(value, mean, .before = 2, .after = 2), 
         .by = type) |>
  filter(value_trend %in% c(min(value_trend), max(value_trend)))

# Completion % in Chomedey/Sainte-Dorothée
comp_pct <- 
  completions_by_type |> 
  filter(!is.na(zone), type == "All", year >= 2014, year <= 2018) |> 
  summarize(val = sum(value[zone == "Chomedey/Sainte-Dorothée"]) / sum(value)) |> 
  pull(val) |> 
  scales::percent(0.1)

# Completion % for rental
rent_pct <- 
  completions_by_market |> 
  filter(is.na(zone), market != "All", year >= 2015) |> 
  summarize(val = sum(value[market == "Rental"]) / sum(value)) |> 
  pull(val) |> 
  scales::percent(0.1)

```

_Data source: CMHC Starts and Completions Survey._

Since 1990, the number of annual housing completions in Laval has remained relatively level, albeit with high levels of year-to-year fluctuation. The five-year period with the most completions has been `r paste0(min_max_vals$year[2] - 2, "-", min_max_vals$year[2] + 2)`, when there was an average of `r min_max_vals$value_trend[2]` housing units completed each year. By contrast, in the period `r paste0(min_max_vals$year[1] - 2, "-", min_max_vals$year[1] + 2)` only `r min_max_vals$value_trend[1]` housing units were completed on average each year.

```{r plot_6_1_1_overall}
plot_6_1_1_overall
```

```{r table_6_1_1_five_year}
table_6_1_1_five_year
```

Prior to 2000, most newly constructed housing in Laval comprised single-family homes, with some semi-detached homes also being constructed particularly in the early 1990s. However, since 2000, there has been a substantial shift away from single-detached homes and towards apartments. Annual single-family housing completions peaked in 2003 and have been declining every year since (albeit with minor fluctuations). By contrast, since 2000 apartment completions have surged, with a five-year moving average demonstrating ongoing increases in the total number of apartments entering the market each year.

There have been no meaningful semi-detached housing completions since the early 1990s, and row houses have consistently been a negligible portion of total completions.

```{r plot_6_1_1_type}
plot_6_1_1_type
```

```{r plot_6_1_1_type_facet}
plot_6_1_1_type_facet
```

The same basic trends are evident looking at housing completions by intended market instead of dwelling type. Homeowner housing completions have steadily declined since a recent peak in the early 2000s. Rental and condominium completions are much more volatile—reflecting a smaller number of large projects. Both increased from near-zero levels in the 1990s. In the case of condominiums, completions began rising in the late 2000s, peaked in 2010, and have since been declining. Rental completions, by contrast, began increasing in the late 2000s and have been increasing ever since. `r rent_pct` of all Laval housing completions since 2015 have been intended for the rental market. There have been no co-operative housing completions in Laval in the period of time that CMHC has tracked this information.

```{r plot_6_1_1_market}
plot_6_1_1_market
```

```{r plot_6_1_1_market_facet}
plot_6_1_1_market_facet
```


Both in absolute and per-dwelling terms, housing development has been concentrated in the southern Chomedey/Sainte-Dorothée region of Laval. This pattern has remained relatively consistent 2009, although was particularly extreme in the 2014-2018 period, where `r comp_pct` of all new housing completions were in the Chomedey/Sainte-Dorothée zone.

```{r map_6_1_1_annual}
map_6_1_1_annual
```


# 6.1.2 Mises en chantier par typologie

```{r s_6_1_2, include = FALSE}

# Min and max 5-year avg
starts_maxima <- 
  starts_by_type |> 
  filter(is.na(zone)) |>
  filter(type == "All") |> 
  mutate(value_trend = slider::slide_dbl(value, mean, .before = 2, .after = 2), 
         .by = type) |> 
  mutate(
    max = slider::slide_lgl(
      value_trend, \(x) all(x[3] >= x, na.rm = TRUE), .before = 2, .after = 2),
    min = slider::slide_lgl(
      value_trend, \(x) all(x[3] <= x, na.rm = TRUE), .before = 2, .after = 2)) |> 
  filter(max | min) |> 
  mutate(max = if_else(year == min(year), TRUE, max),
         min = if_else(year == min(year), FALSE, min),
         max = if_else(year == max(year), FALSE, max),
         min = if_else(year == max(year), TRUE, min)) |> 
  mutate(value_trend = scales::comma(value_trend, 1)) |> 
  filter(max)

starts_minima <- 
  starts_by_type |> 
  filter(is.na(zone)) |>
  filter(type == "All") |> 
  mutate(value_trend = slider::slide_dbl(value, mean, .before = 2, .after = 2), 
         .by = type) |> 
  mutate(
    max = slider::slide_lgl(
      value_trend, \(x) all(x[3] >= x, na.rm = TRUE), .before = 2, .after = 2),
    min = slider::slide_lgl(
      value_trend, \(x) all(x[3] <= x, na.rm = TRUE), .before = 2, .after = 2)) |> 
  filter(max | min) |> 
  mutate(max = if_else(year == min(year), TRUE, max),
         min = if_else(year == min(year), FALSE, min),
         max = if_else(year == max(year), FALSE, max),
         min = if_else(year == max(year), TRUE, min)) |> 
  mutate(value_trend = scales::comma(value_trend, 1)) |> 
  filter(min)

max_years <- paste(paste(starts_maxima$year[1:2], collapse = ", "), starts_maxima$year[3], sep = " and ")
max_vals <- paste(paste(starts_maxima$value_trend[1:2], collapse = ", "), starts_maxima$value_trend[3], sep = " and ")

min_years <- paste(paste(starts_minima$year[1:2], collapse = ", "), starts_minima$year[3], sep = " and ")
min_vals <- paste(paste(starts_minima$value_trend[1:2], collapse = ", "), starts_minima$value_trend[3], sep = " and ")

# Start % in Chomedey/Sainte-Dorothée
start_pct <- 
  starts_by_type |> 
  filter(!is.na(zone), type == "All", year >= 2014, year <= 2018) |> 
  summarize(val = sum(value[zone == "Chomedey/Sainte-Dorothée"]) / sum(value)) |> 
  pull(val) |> 
  scales::percent(0.1)

# Starts in the last three years
starts_21_23 <- 
  starts_by_type |> 
  filter(is.na(zone), type == "All", year >= 2021) |> 
  pull(value) |> 
  sum()

```

_Data source: CMHC Starts and Completions Survey._

Since 1990, housing starts in Laval have varied cyclically, with three distinct peaks and troughs. Rolling five-year averages of housing starts peaked in `r max_years` at `r max_vals` respectively, and bottomed out at `r min_years` at `r min_vals` respectively. (However, although 2023 saw a recent low in housing starts, it is likely that the five-year average will continue to decline for the next several years, given the fact that only `r scales::comma(starts_minima$value[3], 1)`) housing units were initiated in 2023.

```{r plot_6_1_2_overall}
plot_6_1_2_overall
```

```{r table_6_1_2_five_year}
table_6_1_2_five_year
```

Prior to 2000, most newly initated housing projects in Laval comprised single-family homes, with some semi-detached homes also being constructed particularly in the early 1990s. However, since 2000, there has been a substantial shift away from single-detached homes and towards apartments. Annual single-family housing starts peaked in 2003 and have been declining every year since (albeit with minor fluctuations). By contrast, since 2000 apartment starts have surged. On the one hand, apartment projects have been steadily increasing as a proportion of total starts. Every year since 2005, apartment projects have been a majority of all housing starts, and every year since 2015 they have been three quarters or more of all starts. On the other hand, the absolute number of housing starts—apartments included—has been declining. Over the last three years (2021-2023), there have only been `r starts_21_23` housing units initiated in Laval. The implication is that housing completions will soon catch up to this trend, and that very few housing units will be added to the City's total housing stock in the next few years. 

```{r plot_6_1_2_type_facet}
plot_6_1_2_type_facet
```

```{r plot_6_1_2_type_apart}
plot_6_1_2_type_apart
```

In both absolute and per-dwelling terms, there has been a progressive shift in the location of new housing starts in Laval. From 2009-2013, the highest number of absolute housing starts were found in the Chomedey/Sainte-Dorothée zone; in proportional terms this was also the case, although there were still relatively robust per-dwelling housing starts throughout most of the city. In the 2014-2018 period, nearly the entirety of housing starts were located in Chomedey/Sainte-Dorothée, and even in proportional terms most other parts of the city saw very little new housing development. In the last five years, by contrast, although Chomedey/Sainte-Dorothée once again saw the greatest number of total new housing starts, it no longer received the highest number of new housing starts proportional to existing dwellings. Instead, it was Laval-Ouest/Fabreville/Ste-Rose that saw the greatest proportional amount of new housing development.

```{r map_6_1_2_annual}
map_6_1_2_annual
```

# 6.1.3 Mises en chantier par mode d'occupation

```{r s_6_1_3, include = FALSE}

# Min and max 5-year avg
start_market_min_max <- 
  starts_by_market |> 
  filter(is.na(zone)) |>
  filter(market != "All", market != "Unknown", market != "Co-Op") |> 
  mutate(value_trend = slider::slide_dbl(value, mean, .before = 2, .after = 2), 
         .by = market) |> 
  filter(year >= 1992) |> 
  summarize(
    owner_max = max(value_trend[market == "Homeowner"]),
    owner_year = year[value_trend == max(value_trend[market == "Homeowner"])],
    condo_max = max(value_trend[market == "Condo"]),
    condo_year = year[value_trend == max(value_trend[market == "Condo"])],
    rental_max = max(value_trend[market == "Rental"]),
    rental_year = year[value_trend == max(value_trend[market == "Rental"])])
```

_Data source: CMHC Starts and Completions Survey._

Housing starts display similar patterns when broken down by intended market. High rates of new housing projects intended for the homeowner market in the 1990s and 2000s have given way first to condominium starts and more recently to rental starts. Looking at five-year moving averages, homeowner starts peaked in `r paste0(start_market_min_max$owner_year - 2, "-", start_market_min_max$owner_year + 2)` at `r start_market_min_max$owner_max` per year. Condominium starts peaked in `r paste0(start_market_min_max$condo_year - 2, "-", start_market_min_max$condo_year + 2)` at `r start_market_min_max$condo_max` per year. And rental starts peaked in `r paste0(start_market_min_max$rental_year - 2, "-", start_market_min_max$rental_year + 2)` at `r start_market_min_max$rental_max` per year.

```{r table_6_1_3_five_year}
table_6_1_3_five_year
```

```{r plot_6_1_3_market}
plot_6_1_3_market
```

```{r plot_6_1_3_market_facet}
plot_6_1_3_market_facet
```

Mirroring the shift discussed in 6.1.2 towards apartment housing starts, sine 2010 an increasingly high percentage of annual housing starts have been intended for the rental market, albeit with a high level of volatility.

```{r plot_6_1_3_market_rental}
plot_6_1_3_market_rental
```

# 6.1.4 Mises en chantier par nombre de chambres à coucher

CMHC does not make data available about housing starts broken down by bedroom count.

# 6.1.5 Prix des logements neufs

```{r s_6_1_5, include = FALSE}

# 80-20 ratios
ratios <- 
  new_prices |> 
  filter(year %in% c(1998, 2020)) |> 
  mutate(ratio_80_20 = ratio_80_20 - 1) |> 
  pull(ratio_80_20) |> 
  scales::percent()

```

_Data source: CMHC Market Absorption Survey_

CMHC collects data on the selling prices of newly absorbed homeowner and condominium units. According to this data, Laval has seen remarkably consistent annual increases in new housing prices since the 1990s. In 1998, the average selling price of a newly absorbed unit was `r scales::dollar(new_prices$average[1])`, while in 2023 it was `r scales::dollar(new_prices$average[26])`--an increase of `r scales::percent(new_prices$average[26] / new_prices$average[1] - 1)`. Over time, the high-end of the market has seen faster proportionate price increases than the low end of the market. In 1998, the 80th-percentile newly absorbed unit sold for `r ratios[1]` more than the 20th-percentile unit, while in 2020 it was `r ratios[2]`. (The subsequent three years have seen this ratio decline again, but the volatility of the trend makes it too soon to know if that reflects a substantive shift in market conditions.)

```{r plot_6_1_5_percentiles}
plot_6_1_5_percentiles
```

At the same time, the number of absorbed units being counted towards these prices has declined sharply since the early 2000s. Since the CMHC only collects data on homeowner and condominium units, and an increasingly large percentage of new housing in Laval is renter housing, the price of newly absorbed units is an increasingly imprecise metric for measuring housing costs.

```{r plot_6_1_5_units}
plot_6_1_5_units
```

# 6.1.7 Mises en chantier selon le prix par nombre de chambres à coucher

CMHC does not make data available about housing starts broken down by price or bedroom count.
