---
title: "6.1 À but lucratif (abordable, prix du marché)"
author: "Curbcut"
output: html_document
---

```{r setup_6_1, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
qs::qload(here::here("data", "section_6_1.qsm"))
library(tidyverse)
library(gt)
library(patchwork)

```

# 6.1.1 Dév rés récent (répartition territoriale par type de logement et typologie)

```{r s_6_1_1, include = FALSE}

# Min and max 5-year avg
min_max_vals <- 
  completions_by_type |> 
  filter(is.na(zone)) |>
  filter(type == "All") |> 
  mutate(value_trend = slider::slide_dbl(value, mean, .before = 2, .after = 2), 
         .by = type) |>
  filter(value_trend %in% c(min(value_trend), max(value_trend)))

# Completion % in Chomedey/Sainte-Dorothée
comp_pct <- 
  completions_by_type |> 
  filter(!is.na(zone), type == "All", year >= 2014, year <= 2018) |> 
  summarize(val = sum(value[zone == "Chomedey/Sainte-Dorothée"]) / sum(value)) |> 
  pull(val) |> 
  scales::percent(0.1)

```

Since 1990, the number of annual housing completions in Laval has remained relatively level, albeit with high levels of year-to-year fluctuation. The five-year period with the most completions has been `r paste0(min_max_vals$year[2] - 2, "-", min_max_vals$year[2] - 2)`, when there was an average of `r min_max_vals$value_trend[2]` housing units completed each year. By contrast, in the period `r paste0(min_max_vals$year[1] - 2, "-", min_max_vals$year[1] - 2)` only `r min_max_vals$value_trend[1]` housing units were completed on average each year.

```{r plot_6_1_1_overall}
plot_6_1_1_overall
```

```{r table_6_1_1_five_year}
table_6_1_1_five_year
```

Prior to 2000, most newly constructed housing in Laval comprised single-family homes, with some semi-detached homes also being constructed particularly in the early 1990s. However, since 2000, there has been a substantial shift away from single-detached homes and towards apartments. Annual single-family housing completions peaked in 2003 and have been declining every year since (albeit with minor fluctuations). By contrast, since 2000 apartment completions have surged, with a five-year moving average demonstrating ongoing increases in the total number of apartments entering the market each year.

There have been no meaningful semi-detached housing completions since the early 1990s, and row houses have consistenly been a negligible portion of total completions.

```{r plot_6_1_1_type}
plot_6_1_1_type
```

```{r plot_6_1_1_type_facet}
plot_6_1_1_type_facet
```

Both in absolute and per-dwelling terms, housing development has been concentrated in the southern Chomedey/Sainte-Dorothée region of Laval. This pattern has remained relatively consistent 2009, although was particularly extreme in the 2014-2018 period, where `r comp_pct` of all new housing completions were in the Chomedey/Sainte-Dorothée zone.

```{r map_6_1_1_annual}
map_6_1_1_annual
```




# 6.1.2 Mises en chantier par typologie

# 6.1.3 Mises en chantier par mode d'occupation

# 6.1.4 Mises en chantier par nombre de chambres à coucher

# 6.1.5 Prix des logements neufs

# 6.1.7 Mises en chantier selon le prix par nombre de chambres à coucher

# 6.1.12 Valeur foncière (si possible Ventilé par typologie et au pied carré)
