---
title: "6.1 À but lucratif (abordable, prix du marché)"
author: "Curbcut"
output: html_document
---

```{r setup_6_1, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
qs::qload(here::here("data", "section_6_1.qsm"))
library(tidyverse)
library(gt)
library(patchwork)

```

# 6.1.1 Développement résidentiel récent (répartition territoriale par type de logement et typologie)

```{r s_6_1_1, include = FALSE}

# Min and max 5-year avg
min_max_vals <- 
  completions_by_type |> 
  filter(is.na(zone)) |>
  filter(type == "All") |> 
  mutate(value_trend = slider::slide_dbl(value, mean, .before = 2, .after = 2), 
         .by = type) |>
  filter(value_trend %in% c(min(value_trend), max(value_trend)))

# Completion % in Chomedey/Sainte-Dorothée
comp_pct <- 
  completions_by_type |> 
  filter(!is.na(zone), type == "All", year >= 2014, year <= 2018) |> 
  summarize(val = sum(value[zone == "Chomedey/Sainte-Dorothée"]) / sum(value)) |> 
  pull(val) |> 
  scales::percent(0.1)

# Completion % for rental
rent_pct <- 
  completions_by_market |> 
  filter(is.na(zone), market != "All", year >= 2015) |> 
  summarize(val = sum(value[market == "Rental"]) / sum(value)) |> 
  pull(val) |> 
  scales::percent(0.1)

```

_Data source: CMHC Starts and Completions Survey._

Since 1990, the number of annual housing completions in Laval has remained relatively level, albeit with high levels of year-to-year fluctuation. The five-year period with the most completions has been `r paste0(min_max_vals$year[2] - 2, "-", min_max_vals$year[2] + 2)`, when there was an average of `r min_max_vals$value_trend[2]` housing units completed each year. By contrast, in the period `r paste0(min_max_vals$year[1] - 2, "-", min_max_vals$year[1] + 2)` only `r min_max_vals$value_trend[1]` housing units were completed on average each year.

```{r plot_6_1_1_overall}
plot_6_1_1_overall
```

```{r table_6_1_1_five_year}
table_6_1_1_five_year
```

Prior to 2000, most newly constructed housing in Laval comprised single-family homes, with some semi-detached homes also being constructed particularly in the early 1990s. However, since 2000, there has been a substantial shift away from single-detached homes and towards apartments. Annual single-family housing completions peaked in 2003 and have been declining every year since (albeit with minor fluctuations). By contrast, since 2000 apartment completions have surged, with a five-year moving average demonstrating ongoing increases in the total number of apartments entering the market each year.

There have been no meaningful semi-detached housing completions since the early 1990s, and row houses have consistently been a negligible portion of total completions.

```{r plot_6_1_1_type}
plot_6_1_1_type
```

```{r plot_6_1_1_type_facet}
plot_6_1_1_type_facet
```

The same basic trends are evident looking at housing completions by intended market instead of dwelling type. Homeowner housing completions have steadily declined since a recent peak in the early 2000s. Rental and condominium completions are much more volatile—reflecting a smaller number of large projects. Both increased from near-zero levels in the 1990s. In the case of condominiums, completions began rising in the late 2000s, peaked in 2010, and have since been declining. Rental completions, by contrast, began increasing in the late 2000s and have been increasing ever since. `r rent_pct` of all Laval housing completions since 2015 have been intended for the rental market. There have been no co-operative housing completions in Laval in the period of time that CMHC has tracked this information.

```{r plot_6_1_1_market}
plot_6_1_1_market
```

```{r plot_6_1_1_market_facet}
plot_6_1_1_market_facet
```


Both in absolute and per-dwelling terms, housing development has been concentrated in the southern Chomedey/Sainte-Dorothée region of Laval. This pattern has remained relatively consistent 2009, although was particularly extreme in the 2014-2018 period, where `r comp_pct` of all new housing completions were in the Chomedey/Sainte-Dorothée zone.

```{r map_6_1_1_annual}
map_6_1_1_annual
```


# 6.1.2 Mises en chantier par typologie

```{r s_6_1_2, include = FALSE}

# Min and max 5-year avg
starts_maxima <- 
  starts_by_type |> 
  filter(is.na(zone)) |>
  filter(type == "All") |> 
  mutate(value_trend = slider::slide_dbl(value, mean, .before = 2, .after = 2), 
         .by = type) |> 
  mutate(
    max = slider::slide_lgl(
      value_trend, \(x) all(x[3] >= x, na.rm = TRUE), .before = 2, .after = 2),
    min = slider::slide_lgl(
      value_trend, \(x) all(x[3] <= x, na.rm = TRUE), .before = 2, .after = 2)) |> 
  filter(max | min) |> 
  mutate(max = if_else(year == min(year), TRUE, max),
         min = if_else(year == min(year), FALSE, min),
         max = if_else(year == max(year), FALSE, max),
         min = if_else(year == max(year), TRUE, min)) |> 
  mutate(value_trend = scales::comma(value_trend, 1)) |> 
  filter(max)

starts_minima <- 
  starts_by_type |> 
  filter(is.na(zone)) |>
  filter(type == "All") |> 
  mutate(value_trend = slider::slide_dbl(value, mean, .before = 2, .after = 2), 
         .by = type) |> 
  mutate(
    max = slider::slide_lgl(
      value_trend, \(x) all(x[3] >= x, na.rm = TRUE), .before = 2, .after = 2),
    min = slider::slide_lgl(
      value_trend, \(x) all(x[3] <= x, na.rm = TRUE), .before = 2, .after = 2)) |> 
  filter(max | min) |> 
  mutate(max = if_else(year == min(year), TRUE, max),
         min = if_else(year == min(year), FALSE, min),
         max = if_else(year == max(year), FALSE, max),
         min = if_else(year == max(year), TRUE, min)) |> 
  mutate(value_trend = scales::comma(value_trend, 1)) |> 
  filter(min)

max_years <- paste(paste(starts_maxima$year[1:2], collapse = ", "), starts_maxima$year[3], sep = " and ")
max_vals <- paste(paste(starts_maxima$value_trend[1:2], collapse = ", "), starts_maxima$value_trend[3], sep = " and ")

min_years <- paste(paste(starts_minima$year[1:2], collapse = ", "), starts_minima$year[3], sep = " and ")
min_vals <- paste(paste(starts_minima$value_trend[1:2], collapse = ", "), starts_minima$value_trend[3], sep = " and ")

# Start % in Chomedey/Sainte-Dorothée
start_pct <- 
  starts_by_type |> 
  filter(!is.na(zone), type == "All", year >= 2014, year <= 2018) |> 
  summarize(val = sum(value[zone == "Chomedey/Sainte-Dorothée"]) / sum(value)) |> 
  pull(val) |> 
  scales::percent(0.1)

# Starts in the last three years
starts_21_23 <- 
  starts_by_type |> 
  filter(is.na(zone), type == "All", year >= 2021) |> 
  pull(value) |> 
  sum()

```

_Data source: CMHC Starts and Completions Survey._

Since 1990, housing starts in Laval have varied cyclically, with three distinct peaks and troughs. Rolling five-year averages of housing starts peaked in `r max_years` at `r max_vals` respectively, and bottomed out at `r min_years` at `r min_vals` respectively. (However, although 2023 saw a recent low in housing starts, it is likely that the five-year average will continue to decline for the next several years, given the fact that only `r scales::comma(starts_minima$value[3], 1)`) housing units were initiated in 2023.

```{r plot_6_1_2_overall}
plot_6_1_2_overall
```

```{r table_6_1_2_five_year}
table_6_1_2_five_year
```

Avant 2000, la majorité des logements nouvellement construits à Laval étaient des maisons individuelles, bien que quelques maisons jumelées aient également été construites, surtout au début des années 1990. Cependant, depuis 2000, les maisons individuelles ont été largement abandonnées au profit des appartements. Le nombre de logements unifamiliaux achevés a atteint son apogée en 2002, avant de diminuer chaque année par la suite, avec quelques fluctuations mineures.  En revanche, depuis 2000, le nombre d'appartements achevés a fortement augmenté. On the one hand, apartment starts have followed the same broad trends as overall housing starts, with a peak in the mid-2000s and another peak in 2020. On the other hand, driven by a collapse in non-apartment housing starts, apartment projects have been steadily increasing as a proportion of total starts. Every year since 2005, apartment projects have been a majority of all housing starts, and every year since 2015 they have been three quarters or more of all starts. 

```{r plot_6_1_2_type_facet}
plot_6_1_2_type_facet
```

```{r plot_6_1_2_type_apart}
plot_6_1_2_type_apart
```

In both absolute and per-dwelling terms, new housing starts have consistently been highest in the southern part of the city—particularly in the Chomedey/Sainte-Dorothée zone—and lowest in the eastern part of the island.

```{r map_6_1_2_annual}
map_6_1_2_annual
```

# 6.1.3 Mises en chantier par mode d'occupation

```{r s_6_1_3, include = FALSE}

# Min and max 5-year avg
start_market_min_max <- 
  starts_by_market |> 
  filter(is.na(zone)) |>
  filter(market != "All", market != "Unknown", market != "Co-Op") |> 
  mutate(value_trend = slider::slide_dbl(value, mean, .before = 2, .after = 2), 
         .by = market) |> 
  filter(year >= 1992) |> 
  summarize(
    owner_max = scales::comma(max(value_trend[market == "Homeowner"]), 1),
    owner_year = year[value_trend == max(value_trend[market == "Homeowner"])],
    condo_max = scales::comma(max(value_trend[market == "Condo"]), 1),
    condo_year = year[value_trend == max(value_trend[market == "Condo"])],
    rental_max = scales::comma(max(value_trend[market == "Rental"]), 1),
    rental_year = year[value_trend == max(value_trend[market == "Rental"])])
```

_Data source: CMHC Starts and Completions Survey._

Les mises en chantier montrent des tendances similaires lorsqu’elles sont ventilées par marché cible. Les taux élevés de nouveaux projets de logements destinés à l’accession à la propriété dans les années 1990 et 2000 ont progressivement laissé place aux mises en chantier de logements en copropriété et, plus récemment, à celles de logements locatifs. Looking at five-year moving averages, homeowner starts peaked in `r paste0(start_market_min_max$owner_year - 2, "-", start_market_min_max$owner_year + 2)` at `r start_market_min_max$owner_max` per year. Condominium starts peaked in `r paste0(start_market_min_max$condo_year - 2, "-", start_market_min_max$condo_year + 2)` at `r start_market_min_max$condo_max` per year. And rental starts peaked in `r paste0(start_market_min_max$rental_year - 2, "-", start_market_min_max$rental_year + 2)` at `r start_market_min_max$rental_max` per year.

```{r table_6_1_3_five_year}
table_6_1_3_five_year
```

```{r plot_6_1_3_market_facet}
plot_6_1_3_market_facet
```

Mirroring the shift discussed in 6.1.2 towards apartment housing starts, sine 2010 an increasingly high percentage of annual housing starts have been intended for the rental market, albeit with a high level of volatility. This trend largely reflects the interaction between a highly cyclical rental construction sector (with peaks in starts in 2007-2008 and 2020), a long-term decline in new starts in the homeownership sector, and the 2000s growth and then 2010s decline in condominium starts.

```{r plot_6_1_3_market_rental}
plot_6_1_3_market_rental
```

# 6.1.4 Mises en chantier par nombre de chambres à coucher

CMHC does not make data available about housing starts broken down by bedroom count.

# 6.1.5 Prix des logements neufs

```{r s_6_1_5, include = FALSE}

# 80-20 ratios
ratios <- 
  new_prices |> 
  filter(year %in% c(1998, 2020)) |> 
  mutate(ratio_80_20 = ratio_80_20 - 1) |> 
  pull(ratio_80_20) |> 
  scales::percent()

```

_Data source: CMHC Market Absorption Survey_

CMHC collects data on the selling prices of newly absorbed homeowner and condominium units. According to this data, Laval has seen remarkably consistent annual increases in new housing prices since the 1990s. In 1998, the average selling price of a newly absorbed unit was `r scales::dollar(new_prices$average[1])`, while in 2023 it was `r scales::dollar(new_prices$average[26])`--an increase of `r scales::percent(new_prices$average[26] / new_prices$average[1] - 1)`. Over time, the high-end of the market has seen faster proportionate price increases than the low end of the market. In 1998, the 80th-percentile newly absorbed unit sold for `r ratios[1]` more than the 20th-percentile unit, while in 2020 it was `r ratios[2]`. (The subsequent three years have seen this ratio decline again, but the volatility of the trend makes it too soon to know if that reflects a substantive shift in market conditions.)

```{r plot_6_1_5_percentiles}
plot_6_1_5_percentiles
```

At the same time, the number of absorbed units being counted towards these prices has declined sharply since the early 2000s. Since the CMHC only collects data on homeowner and condominium units, and an increasingly large percentage of new housing in Laval is renter housing, the price of newly absorbed units is an increasingly imprecise metric for measuring housing costs.

```{r plot_6_1_5_units}
plot_6_1_5_units
```

# 6.1.7 Mises en chantier selon le prix par nombre de chambres à coucher

CMHC does not make data available about housing starts broken down by price or bedroom count.
