---
title: "6.1 À but lucratif (abordable, prix du marché)"
author: "Curbcut"
output: html_document
---

```{r setup_6_1, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
qs::qload(here::here("data", "section_6_1.qsm"))
library(tidyverse)

```

# 6.1.1 Dév rés récent (répartition territoriale par type de logement et typologie)

```{r s_6_1_1, include = FALSE}

# Min and max 5-year avg
min_max_vals <- 
  completions_by_type |> 
  filter(is.na(zone)) |>
  filter(type == "All") |> 
  mutate(value_trend = slider::slide_dbl(value, mean, .before = 2, .after = 2), 
         .by = type) |>
  filter(value_trend %in% c(min(value_trend), max(value_trend)))


```

Since 1990, the number of annual housing completions in Laval has remained relatively level, albeit with high levels of year-to-year fluctuation. The five-year period with the most completions has been `r paste0(min_max_vals$year[2] - 2, "-", min_max_vals$year[2] - 2)`, when there was an average of `r min_max_vals$value_trend[2]` housing units completed each year. By contrast, in the period `r paste0(min_max_vals$year[1] - 2, "-", min_max_vals$year[1] - 2)` only `r min_max_vals$value_trend[1]` housing units were completed on average each year.


# 6.1.2 Mises en chantier par typologie

# 6.1.3 Mises en chantier par mode d'occupation

# 6.1.4 Mises en chantier par nombre de chambres à coucher

# 6.1.5 Prix des logements neufs

# 6.1.7 Mises en chantier selon le prix par nombre de chambres à coucher

# 6.1.12 Valeur foncière (si possible Ventilé par typologie et au pied carré)
